Viscoelastic Kelvin-Helmholtz instability

### 2014-05-15 paper/paper.tex
updated paper a bit

%	paper.tex	Toby Searle
%	Last modified: Tue 29 Oct 10:24:24 2013
%	using jfm template

@@

For a long time it was thought that the elasticity of a viscoelastic fluid only served to dampen instability. It was observed that the drag in fluid could be reduced by the addition of small amounts of viscoelastic fluid. This drag reduction has been much studied over the last 70 years. However, the studies above give reason to believe that viscoelastic turbulence is present at very low Reynolds number, where it is driven by the elasticity of the polymeric fluid rather than its inertia.

In 1997 Fabian Waleffe identified a Newtonian self sustaining process in plane Couette flow \cite{Waleffe97}. Streamwise rolls redistribute the streamwise velocity into a streaky flow. This streaky flow is unstable through a Kelvin-Helmholtz instability creating a three dimensional perturbation which in turn re-energises the original streamwise rolls. This exact solution to the Navier-Stokes equations is thought to be a component of the transition to turbulence. In 1998 Waleffe constructed a bifurcation diagram for this exact solution \cite{Waleffe98} containing what looks like a bifurcation from infinity. This is the behaviour expected of the transition to turbulence in Couette flow.

Using this work in Newtonian fluid dynamics as a template for the structure of purely elastic turbulence reveals a similar pattern of streaks and, hopefully, a similar self-sustaining process. An important step towards explaining this process in purely elastic flows is establishing the mechanism by which the streaks become unstable. Presumably, this will also be a Kelvin-Helmholtz instability as the streaks shear with the rest of the fluid. The Kelvin-Helmholtz instability is also often integral to the transition to turbulence in Newtonian fluids. It is hoped that by finding a purely elastic version of this instability, we might be able to probe one of the important mechanisms for the transition to purely elastic turbulence. To explore this mechanism in the purely elastic regime we have constructed a simple model system of a shearing viscoelastic fluid.

The key dimensionless number for turbulence without inertia in Viscoelastic flows is the Weissenberg number, which is the ratio of the normal stress to the shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma}$. For small shear rates, this is approximately equivalent to $De$, the Deborah number, so that $Wi \approx De = \lambda \dot{\gamma}$ where $\dot{\gamma}$ is the shear rate.

We present a linear stability analysis of the Kelvin-Helmholtz instability in the purely elastic regime using direct numerical simulation of both the Oldroyd-B and FENE-CR constitutive models for a polymeric fluid. The purely elastic fluid is found to be unstable at sufficiently high numbers of the dimensionless parameter, the Weissenberg number. The growth rate of the largest growing eigenmode of the instability is found to increase with increasing Weissenberg number for low Reynold's number. This behaviour is consistent across both the Oldroyd-B and FENE-CR models.

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low Reynolds number $\Rey < 0.01$ and a small ratio between the solvent and total viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid, where the polymer contribution to the dynamics is higher than the solvent contribution. The Oldroyd-B Navier stokes and constitutive equations are then:
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$ as the total stress in the polymeric fluid. The FENE-CR viscoelastic fluid has the advantage of being similarly easy to simulate, but also including a finite extensibility for the polymers. This leads to a stress which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear spring force:

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_vary_Re}
	\caption{}
	\label{fig:inf_vary_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Wi}
	\caption{}
	\label{fig:dispersions_low_Wi}
    \end{subfigure}
    \caption{a) Free shear version of the instability. Plot of the maximum growth rate against Reynold's number with $\Wi=50$. b) Dispersion relations at various Reynold's numbers to accompany the trend in figure a).}
\end{figure}

@@

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{psi_low_delta}
    \caption{}
    \label{fig:psi_low_delta}
    \end{subfigure}
    \vspace{1 mm}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{txx_low_delta}
    \caption{}
    \label{fig:stress_low_delta}
    \end{subfigure}
    \caption{a) Streamfunction of the largest eigenfunction of the instability when $\Rey = 0.1$, $\Wi =100$, $\Delta = 0.005$. Red is positive (streamlines of the flow point clockwise around the contours), Blue is negative (streamlines of the flow point anticlockwise around the contours) b) xx stress for the same instability.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{KH_vary_delta}
    \caption{Dispersion relations for various values of $\Delta$ at $\Rey = 0.1$, $\Wi = 100$. The sensitivity to the walls seems only to increase the strength of the instability. Even when the system is insensitive to the walls, at $\Delta= 0.0125$, there is still an instability present.}
    \label{fig:walls_dependence}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{psi_high_delta}
    \caption{Streamfunction of the eigenfunction of the instability at $\Delta = 0.1$, $\Wi = 100$, $\Rey=0.1$. The magnitude is scaled by the xx stress in the centre of the channel.}
    \label{fig:psi_high_delta}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{vary_Re_delta_conv}
    \caption{The blue line is the data from the stretched code and the green line is the $\Delta = 0.1$ data. By $\Rey = 1$ it is clear that the Reynold's number is sufficient such that the walled system behaves as though it were insensitive to the walls.}
    \label{fig:vary_Re_delta_conv}
\end{figure}

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{a) The change in the height of the dispersion relation as $\Wi$ is increased at $\Rey = 0$ with $\Delta = 0.1$. This instability must be entirely elastic in nature. b) The change in the height of the dispersion relation as the Reynolds number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still present for large polymer concentrations, consistent with purely elastic turbulence with $\Delta = 0.1$.  c) The value of the Weissenberg number at which the velocity profile becomes unstable against $\beta$ at $\Delta = 0.1$. The onset of the purely elastic instability moves to higher $Wi$ as $\beta$ increases.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{FENE_purely_elastic}
    \caption{The FENE-CR model fluid at $\Rey = 0$  when $\Delta = 0.1$. Results are consistent with figure \ref{fig:KH_purely_elastic}.}
    \label{fig:FENE_low_Re}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{high_Re_vary_delta}
    \caption{ Dispersion relations for various values of $\Delta$ and $\Rey = 1000$, $\Wi = 0.01$. When results are rescaled for delta, The instability is found to be insensitive to the width of the instability relative to the wall at this combination of Reynold's number and Wiessenberg number.}
    \label{fig:delta_dispersion}
\end{figure}

@@

\begin{figure}
%    \showthe\columnwidth %use this to determine size of figures.
    \centering
    \includegraphics[width=0.7\textwidth]{KH_high_Re_vary_Wi}
    \caption{Peak growth rate against Wiessenberg number at $\Rey = 1000$, $\Delta = 0.1$ The height of the dispersion relation is reduced for the Kelvin-Helmholtz instability as the Weissenberg number is increased. The grey dashed line is the height of the dispersion relation for the Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased stability of the base flow and so confirms the presence of drag reduction in this system.}
    \label{fig:KH_drag_reduction}
\end{figure}

>>>>

%	paper.tex	Toby Searle
%	Last modified: Wed 14 May 22:10:55 2014
%	using jfm template

@@

It was long thought that the elasticity of a viscoelastic fluid only served to
dampen instability. This view stems from early results showing that drag ina
turbulent Newtonian flow can be reduced by the addition of small amounts of
viscoelastic fluid (\cite{Toms1977}). Although drag reduction has recieved a
lot of attention in the past the studies above give reason to believe that
there is a potentially more prevalent and useful phenomenon at work in
viscoelastic fluids. It seems that viscoelastic turbulence is present at very
low Reynolds number, where it is driven by the elasticity of the polymeric
fluid rather than its inertia.

In 1997 Fabian Waleffe identified a Newtonian self sustaining process in plane
Couette flow \cite{Waleffe1997}. Streamwise rolls redistribute the streamwise
velocity into a streaky flow. This streaky flow is unstable through a
Kelvin-Helmholtz instability leading to a symmetry bifurcation to a three
dimensional perturbation which in turn re-energises the original streamwise
rolls. This exact solution to the Navier-Stokes equations is thought to be a
component of the transition to turbulence. In 1998 Waleffe constructed a
bifurcation diagram for this exact solution \cite{Waleffe1998} containing what
looks like a bifurcation from infinity. This is the behaviour expected of the
transition to turbulence in Couette flow.

Using this work in Newtonian fluid dynamics as a template for the structure of
purely elastic turbulence reveals a similar pattern of streaks and, hopefully,
a similar self-sustaining process \cite{US!}. An important step towards
explaining this process in purely elastic flows is establishing the mechanism
by which the streaks become unstable. Presumably, this will also be a
Kelvin-Helmholtz instability as the streaks shear with the rest of the fluid.
The Kelvin-Helmholtz instability is also often integral to the transition to
turbulence in Newtonian fluids. It is hoped that by finding a purely elastic
version of this instability, we might be able to probe one of the important
mechanisms for the transition to purely elastic turbulence. To explore this
mechanism in the purely elastic regime we have constructed a simple model
system of a shearing viscoelastic fluid.

The key dimensionless number for turbulence without inertia in viscoelastic
flows is the Weissenberg number, which is the ratio of the normal stress to the
shear stress of the polymer component of the fluid $Wi = \frac{N_{1}}{\sigma}$.
For small shear rates, this is approximately equivalent to $De$, the Deborah
number, so that $Wi \approx De = \lambda \dot{\gamma}$ where $\dot{\gamma}$ is
the shear rate.

We present a linear stability analysis of the Kelvin-Helmholtz instability in
the purely elastic regime using direct numerical simulation of both the
Oldroyd-B and FENE-P constitutive models for a polymeric fluid. The purely
elastic fluid is found to be unstable at sufficiently high numbers of the
dimensionless parameter, the Weissenberg number. The growth rate of the largest
growing eigenmode of the instability is found to increase with increasing
Weissenberg number for low Reynold's number. This behaviour is consistent
across both the Oldroyd-B and FENE-P models.

@@

In order to simulate this flow, we used an Oldroyd-B fluid with very low
Reynolds number $\Rey < 0.01$ and a small ratio between the solvent and total
viscosities $\beta = \frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a
purely elastic fluid, where the polymer contribution to the dynamics is higher
than the solvent contribution. The Oldroyd-B Navier stokes and constitutive
equations are then: 
\begin{align}
    Re \left[ \dt{\mathbf{v}} + \mathbf{v} \cdot \nabla  \mathbf{v} \right] &=
    - \nabla p + \beta \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot
    \mathbf{\tau} \\ \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla
    \mathbf{v}\right)^{T} + \nabla{\mathbf{v}}
\end{align}
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$
as the total stress in the polymeric fluid. The FENE-P viscoelastic fluid has
the advantage of being similarly easy to simulate, but also including a finite
extensibility for the polymers. This leads to a stress which depends on a
conformation tensor ($\mathbf{C}$) for the polymer dumbbells via a non-linear
spring force:

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_vary_Re}
	\caption{}
	\label{fig:inf_vary_Re}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.48\textwidth}
	\centering
	\includegraphics[width=\textwidth]{inf_dispersions_low_Wi}
	\caption{}
	\label{fig:dispersions_low_Wi}
    \end{subfigure}
    \caption{
	a) Free shear version of the instability. Plot of the maximum growth
	rate against Reynold's number with $\Wi=50$. b) Dispersion relations at
	various Reynold's numbers to accompany the trend in figure a).
    } 
    \end{figure}

@@

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{psi_low_delta}
    \caption{}
    \label{fig:psi_low_delta}
    \end{subfigure}
    \vspace{1 mm}
    \begin{subfigure}[b]{\textwidth}
    \centering
    \includegraphics[width=\textwidth]{txx_low_delta}
    \caption{}
    \label{fig:stress_low_delta}
    \end{subfigure}
    \caption{
	a) Streamfunction of the largest eigenfunction of the instability when
	$\Rey = 0.1$, $\Wi =100$, $\Delta = 0.005$. Red is positive
	(streamlines of the flow point clockwise around the contours), Blue is
	negative (streamlines of the flow point anticlockwise around the
	contours) b) xx stress for the same instability.
    }
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{KH_vary_delta}
    \caption{
    Dispersion relations for various values of $\Delta$ at $\Rey = 0.1$, $\Wi =
    100$. The sensitivity to the walls seems only to increase the strength of
    the instability. Even when the system is insensitive to the walls, at
    $\Delta= 0.0125$, there is still an instability present.
    }
    \label{fig:walls_dependence}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{psi_high_delta}
    \caption{
    Streamfunction of the eigenfunction of the instability at $\Delta = 0.1$,
    $\Wi = 100$, $\Rey=0.1$. The magnitude is scaled by the xx stress in the
    centre of the channel.
    }
    \label{fig:psi_high_delta}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{vary_Re_delta_conv}
    \caption{
	The blue line is the data from the stretched code and the green line is
	the $\Delta = 0.1$ data. By $\Rey = 1$ it is clear that the Reynold's
	number is sufficient such that the walled system behaves as though it
	were insensitive to the walls.
    }
    \label{fig:vary_Re_delta_conv}
\end{figure}

@@

\begin{subfigure}[b]{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{KH_onset_beta_Wi}
    \caption{}
    \label{fig:KH_elastic_onset}
\end{subfigure}
\caption{
a) The change in the height of the dispersion relation as $\Wi$ is increased at
$\Rey = 0$ with $\Delta = 0.1$. This instability must be entirely elastic in
nature. b) The change in the height of the dispersion relation as the Reynolds
number is reduced at $\Wi=5$. At low $\Rey$ the instability is clearly still
present for large polymer concentrations, consistent with purely elastic
turbulence with $\Delta = 0.1$.  c) The value of the Weissenberg number at
which the velocity profile becomes unstable against $\beta$ at $\Delta = 0.1$.
The onset of the purely elastic instability moves to higher $Wi$ as $\beta$
increases.
}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{FENE_purely_elastic}
    \caption{
    The FENE-P model fluid at $\Rey = 0$  when $\Delta = 0.1$. Results are
    consistent with figure \ref{fig:KH_purely_elastic}.
    }
    \label{fig:FENE_low_Re}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{high_Re_vary_delta}
    \caption{ 
    Dispersion relations for various values of $\Delta$ and $\Rey = 1000$, $\Wi
    = 0.01$. When results are rescaled for delta, The instability is found to
    be insensitive to the width of the instability relative to the wall at this
    combination of Reynold's number and Wiessenberg number.
    }
    \label{fig:delta_dispersion}
\end{figure}

@@

\begin{figure}
%    \showthe\columnwidth %use this to determine size of figures.
    \centering
    \includegraphics[width=0.7\textwidth]{KH_high_Re_vary_Wi}
    \caption{
	Peak growth rate against Wiessenberg number at $\Rey = 1000$, $\Delta =
	0.1$ The height of the dispersion relation is reduced for the
	Kelvin-Helmholtz instability as the Weissenberg number is increased.
	The grey dashed line is the height of the dispersion relation for the
	Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased
	stability of the base flow and so confirms the presence of drag
	reduction in this system.
    }
    \label{fig:KH_drag_reduction}
\end{figure}
